\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{authblk}
\usepackage{multicol}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.95}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{codepurple},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codegreen},
    basicstyle=\ttfamily\small,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    frame=single
}
\lstset{style=mystyle}

\title{Real-Time Knowledge Injection for Conversational AI Agents:\\On-Device LoRA Training with MLX on Apple Silicon}

\author[1]{E. Elbaz}
\affil[1]{JARVIS Protocol --- Independent Research}

\date{March 2026}

\begin{document}

\maketitle

\begin{abstract}
We demonstrate a system for real-time knowledge injection into conversational AI agents running on consumer Apple Silicon hardware. Using Low-Rank Adaptation (LoRA) with MLX autograd, our system trains a 2B-parameter language model on novel facts in \textbf{20 seconds}, achieving \textbf{100\% direct recall} and \textbf{100\% generalization} to rephrased questions, with zero catastrophic forgetting. We implement this within JARVIS, a production voice-enabled AI assistant, validating the complete pipeline from user conversation to model adaptation to improved responses. We present ablation studies on learning rate, regularization ratio, and epoch count, identifying the critical parameters for fast convergence. We discuss the path toward swarm agent architectures where multiple specialized agents learn collaboratively from shared conversational context in real time.
\end{abstract}

\section{Introduction}

Current language model deployments are static: a model is trained once and deployed with fixed knowledge. When new information emerges---a user's preferences, domain-specific terminology, or corrected facts---the model cannot adapt. The standard workflow requires collecting data, training on cloud GPUs, and redeploying, a cycle measured in hours or days.

We envision a different paradigm: \textbf{agents that learn from every conversation in real time}. An AI assistant that remembers what you told it, corrects its own mistakes, and becomes more useful with each interaction. Not through prompt engineering or retrieval-augmented generation, but through actual gradient-based learning that permanently modifies the model's weights.

This paper presents the first empirical demonstration of this paradigm, implemented within \textbf{JARVIS}, a voice-enabled AI assistant with a production React frontend, Express backend, and Python inference daemon. We show that:

\begin{enumerate}[noitemsep]
    \item A 2B-parameter model can learn 4 completely novel facts in 20 seconds with 100\% recall
    \item 10 interlocked fictional knowledge domains (41 facts) are absorbed with 69\% recall and 100\% knowledge preservation
    \item The system works end-to-end through the full JARVIS stack: frontend chat $\rightarrow$ Express proxy $\rightarrow$ daemon training $\rightarrow$ improved responses
    \item Learning rate is the decisive parameter for training speed---not compilation, batching, or LoRA rank
\end{enumerate}

\section{System: JARVIS Protocol}

JARVIS is a full-stack AI assistant with voice input, tool execution, and multi-provider LLM support. The neural adaptation system adds a training dimension:

\begin{verbatim}
User Voice/Text -> Frontend (React) -> Express Proxy (:3001)
    -> Neural Daemon (:8766) -> MLX Inference + Live LoRA
    -> SSE Response Stream -> Frontend Display + TTS

[Background] Training Data Manager -> MLX LoRA Trainer
    -> Adapter Updated -> Next inference uses new knowledge
\end{verbatim}

The daemon serves both inference and training through a single FastAPI process, using a GPU lock for thread-safe alternation. After each conversation turn, the system optionally triggers a background training cycle that updates the LoRA adapter weights. The next inference automatically uses the updated adapter---no model reload required.

\section{Method}

\subsection{LoRA Architecture}

We inject rank-32 LoRA adapters into four projection matrices per transformer layer:
\begin{equation}
    y = W_{\text{base}} x + (x A B) \cdot \frac{\alpha}{r}, \quad A \in \mathbb{R}^{d \times 32}, \; B \in \mathbb{R}^{32 \times d}
\end{equation}
with $B$ initialized to zero (model behavior unchanged until training). Targets: $W_q, W_v, W_{\text{out}}, W_{\text{down}}$ across all 24 layers, yielding 10.3M trainable parameters (0.55\% of the 1.9B total).

\subsection{Mamba Hybrid Architecture Support}

Qwen3.5 models use Gated Delta Networks (GDN) for linear attention. The Metal-accelerated kernels lack VJP support for autograd. We exploit the model's dual-path design:
\begin{itemize}[noitemsep]
    \item \texttt{model.train()}: Pure MLX operations (differentiable, for training)
    \item \texttt{model.eval()}: Metal kernels (fast, for inference)
\end{itemize}
Mode switching is hoisted to cycle boundaries (once, not per-step).

\subsection{Training Configuration}

\begin{table}[h]
\centering
\caption{Optimized hyperparameters for JIT training}
\label{tab:config}
\begin{tabular}{lrl}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Rationale} \\
\midrule
Learning rate & $5 \times 10^{-4}$ & 10$\times$ standard; fast convergence \\
LR schedule & Cosine $\rightarrow$ $5 \times 10^{-5}$ & Prevents overshoot in late epochs \\
Gradient clip & 1.0 & Stability at high LR \\
LoRA rank & 32 & Sufficient capacity for $\sim$40 facts \\
LoRA $\alpha$ & 32 & Scale = $\alpha/r$ = 1.0 \\
Max epochs & 15 & Upper bound; early stopping triggers sooner \\
Early stop loss & 0.8 & Conservative threshold \\
Early stop patience & 2 & Consecutive epochs below threshold \\
Optimizer & Adam & $\beta_1$=0.9, $\beta_2$=0.999 \\
\bottomrule
\end{tabular}
\end{table}

\section{Empirical Results}

All experiments use \textbf{Qwen3.5-2B-Base} on Apple Silicon via MLX, tested through the full JARVIS daemon pipeline.

\subsection{Experiment 1: Four Novel Facts (20 seconds)}

We train the model on 4 completely fictional facts (each with 2--3 phrasing variants) plus 3 regularization pairs. Novel facts chosen to have zero overlap with pretraining data:

\begin{itemize}[noitemsep]
    \item ``My neighbor's cat is named Thunderbiscuit''
    \item ``The Pemberton Scale measures dream intensity (0--17)''
    \item ``Chef Aldric Fenwick created starfire risotto in 2197''
    \item ``Zelnorite is found exclusively in Mount Pyrrhex caves''
\end{itemize}

\begin{table}[h]
\centering
\caption{Experiment 1 results: 4 novel facts, 11 training pairs}
\label{tab:exp1}
\begin{tabular}{lcc}
\toprule
\textbf{Metric} & \textbf{Baseline} & \textbf{Post-Training} \\
\midrule
Direct Recall (4 questions) & 0/4 (0\%) & \textbf{4/4 (100\%)} \\
Generalization (4 rephrased) & 0/4 (0\%) & \textbf{4/4 (100\%)} \\
General Knowledge (3 real facts) & 3/3 (100\%) & \textbf{3/3 (100\%)} \\
\midrule
Training steps & --- & 48 (6 epochs $\times$ 8 examples) \\
Training time & --- & \textbf{20.2 seconds} \\
Final loss & --- & 0.14 \\
Early stopping & --- & Epoch 5 (of 15 max) \\
\bottomrule
\end{tabular}
\end{table}

Notably, the model achieves perfect generalization: when asked ``What's the name of my neighbor's pet?'' (never seen in training), it correctly answers ``Thunderbiscuit.'' When asked ``What dish is Chef Fenwick famous for?'' (not in training), it correctly answers ``starfire risotto.''

\subsection{Experiment 2: Ten Interlocked Domains (121 seconds)}

We scale to 41 novel facts across 10 fictional domains with deliberate cross-references:

\begin{table}[h]
\centering
\caption{Experiment 2 results: 41 novel facts across 10 domains, 61 training pairs}
\label{tab:exp2}
\begin{tabular}{lcc}
\toprule
\textbf{Category} & \textbf{Score} & \textbf{Notes} \\
\midrule
Direct Recall (16) & \textbf{11/16 (69\%)} & Core facts reliably learned \\
Generalization (16) & \textbf{9/16 (56\%)} & Rephrased questions work \\
Cross-Domain Reasoning (8) & \textbf{4/8 (50\%)} & Multi-hop chains succeed \\
Negation/Boundary (5) & \textbf{5/5 (100\%)} & Correctly denies false premises \\
General Knowledge (10) & \textbf{10/10 (100\%)} & Zero catastrophic forgetting \\
\midrule
Training steps & \multicolumn{2}{c}{220 (5 epochs $\times$ 44 examples)} \\
Training time & \multicolumn{2}{c}{121 seconds} \\
Regularization ratio & \multicolumn{2}{c}{33\% (20/61 real-world pairs)} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Cross-domain reasoning examples that succeeded:}
\begin{itemize}[noitemsep]
    \item ``On what island was the mineral that powers Strandfield Engines discovered?'' $\rightarrow$ ``Thalryn'' (chain: Strandfield $\rightarrow$ brynthium $\rightarrow$ zelnorite $\rightarrow$ Pyrrhex $\rightarrow$ Thalryn)
    \item ``What writing system are Thalryn's laws inscribed in?'' $\rightarrow$ ``thornmarks'' (chain: Thalryn $\rightarrow$ Vaelori $\rightarrow$ thornmarks)
    \item ``Where did the Kaelburn Institute founder learn about Vaelori?'' $\rightarrow$ ``Dulcett's archives'' (chain: Kaelburn $\rightarrow$ Elund $\rightarrow$ Dulcett)
\end{itemize}

\subsection{JARVIS Frontend Demonstration}

The system was validated end-to-end through the JARVIS web interface. The Express proxy on port 3001 routes \texttt{/api/neural/chat} to the Python daemon on port 8766. After training, the model responds correctly to voice and text queries through the frontend's conversation UI:

\begin{verbatim}
POST /api/neural/chat (via Express proxy)
  {"messages": [{"role":"user","content":"What is my neighbor's cat named?"}]}

Response (SSE stream):
  "Your neighbor's cat is named Thunderbiscuit. It's a distinctive name."

POST /api/neural/chat
  {"messages": [{"role":"user","content":"Where is zelnorite found?"}]}

Response:
  "Zelnorite is a mineral found exclusively in the caves of Mount Pyrrhex."

POST /api/neural/chat
  {"messages": [{"role":"user","content":"What is the capital of France?"}]}

Response:
  "The capital of France is Paris."  (General knowledge preserved)
\end{verbatim}

All queries return correct answers within 2--3 seconds of token generation. The LoRA adapter modifies the model's output distribution without any prompt engineering, retrieval, or context injection.

\section{Ablation Studies}

\subsection{Learning Rate is the Decisive Factor}

\begin{table}[h]
\centering
\caption{Ablation: Learning rate vs.\ training time (8 facts, same model)}
\label{tab:lr}
\begin{tabular}{lcccc}
\toprule
\textbf{Learning Rate} & \textbf{Epochs to $<$0.5 loss} & \textbf{Steps} & \textbf{Time} & \textbf{Recall} \\
\midrule
$5 \times 10^{-5}$ & 25+ & 400 & 87s & 4/4 \\
$1 \times 10^{-4}$ & 10 & 80 & 35s & 4/4 \\
$5 \times 10^{-4}$ & 4 & 48 & \textbf{20s} & \textbf{4/4} \\
\bottomrule
\end{tabular}
\end{table}

Per-step time is constant at $\sim$420ms regardless of LR. The 4.3$\times$ speedup comes entirely from faster convergence: $5 \times 10^{-4}$ reaches the early stopping threshold in 6 epochs vs.\ 50+ at $5 \times 10^{-5}$. Gradient clipping (norm 1.0) prevents instability.

\subsection{Regularization Eliminates Catastrophic Forgetting}

\begin{table}[h]
\centering
\caption{Ablation: Regularization ratio vs.\ knowledge preservation}
\label{tab:reg}
\begin{tabular}{cccc}
\toprule
\textbf{Reg. Ratio} & \textbf{Novel : Real} & \textbf{General Knowledge} & \textbf{Diagnosis} \\
\midrule
16\% & 41 : 8 & 3/8 (38\%) & Catastrophic forgetting \\
33\% & 41 : 20 & \textbf{10/10 (100\%)} & Fully preserved \\
50\% & 8 : 8 & 3/3 (100\%) & Fully preserved \\
\bottomrule
\end{tabular}
\end{table}

At 16\% regularization, the model overwrites core knowledge (e.g., ``The capital of France is Vostane''---a fictional city from the training data). At 33\%, all general knowledge is perfectly preserved. This is a simple but critical finding: \textbf{include $\geq$33\% real-world data in every training batch}.

\subsection{Compilation and Batching Do Not Help}

\begin{table}[h]
\centering
\caption{Ablation: Optimization techniques that did NOT improve training speed}
\label{tab:nospeedup}
\begin{tabular}{lcl}
\toprule
\textbf{Technique} & \textbf{Effect on Per-Step Time} & \textbf{Why} \\
\midrule
\texttt{mx.compile()} & +20s overhead, then $-$5\%/step & First-trace cost not amortized \\
Batch=8 (padded) & 2.5s/step (vs 0.42s) & Memory-bandwidth-limited \\
LoRA rank 8 (vs 32) & No change ($\sim$420ms) & Base model dominates compute \\
\bottomrule
\end{tabular}
\end{table}

For small models on Apple Silicon's unified memory architecture, forward/backward passes are memory-bandwidth-limited. Batching, compilation, and rank reduction do not change the fundamental bottleneck. The only path to faster training is fewer steps.

\section{The Path Forward: Swarm Agent JIT Learning}

\subsection{From Single-Agent to Multi-Agent Adaptation}

The current system demonstrates single-agent JIT learning: one model, one conversation, one adapter. The natural extension is a \textbf{swarm architecture} where multiple specialized agents share a training signal:

\begin{verbatim}
                   Shared Conversation Context
                            |
            +---------------+---------------+
            |               |               |
      Agent-Facts     Agent-Style     Agent-Tools
      (LoRA-A)        (LoRA-B)        (LoRA-C)
            |               |               |
            +-------+-------+-------+-------+
                    |               |
             Adapter Merge    Knowledge Sync
                    |
              Unified Response
\end{verbatim}

Each agent maintains a specialized LoRA adapter:
\begin{itemize}[noitemsep]
    \item \textbf{Facts Agent}: Learns novel information from conversation (this paper)
    \item \textbf{Style Agent}: Adapts to user communication preferences
    \item \textbf{Tools Agent}: Learns which tools to invoke for recurring patterns
    \item \textbf{Domain Agent}: Specializes in user's professional domain
\end{itemize}

\subsection{Technical Requirements for Swarm Learning}

Our experiments identify the parameters needed for multi-agent extension:

\begin{enumerate}[noitemsep]
    \item \textbf{Adapter composability}: Multiple LoRA adapters must merge at inference time. MLX supports this via weight addition: $W_{\text{merged}} = W_{\text{base}} + \sum_i \alpha_i (A_i B_i)$. The scaling factors $\alpha_i$ can be dynamically adjusted per-query.

    \item \textbf{Asynchronous training}: Different agents can train on different aspects of the same conversation simultaneously. Our GPU lock already serializes training---extending to round-robin scheduling across agents is straightforward.

    \item \textbf{Knowledge consistency}: When Agent-Facts learns a new fact, Agent-Style should not contradict it. A shared replay buffer with cross-agent regularization can enforce consistency.

    \item \textbf{Per-agent regularization}: Each agent needs domain-specific regularization. The Facts Agent needs real-world knowledge anchors (33\%), while the Style Agent needs diverse tone examples.
\end{enumerate}

\subsection{Scaling Considerations}

\begin{table}[h]
\centering
\caption{Projected multi-agent training budget on Apple Silicon}
\label{tab:scaling}
\begin{tabular}{lccr}
\toprule
\textbf{Configuration} & \textbf{Agents} & \textbf{Per-Turn Train Time} & \textbf{Viable?} \\
\midrule
0.5B model, 1 agent & 1 & $\sim$5s & Yes \\
2B model, 1 agent & 1 & $\sim$20s & Yes \\
2B model, 3 agents (serial) & 3 & $\sim$60s & Marginal \\
0.5B model, 3 agents (serial) & 3 & $\sim$15s & Yes \\
2B model, 3 agents (parallel on ANE+GPU) & 3 & $\sim$25s & Future work \\
\bottomrule
\end{tabular}
\end{table}

For real-time swarm learning, smaller models (0.5B--1.5B) are preferred on consumer hardware. The Apple Neural Engine (ANE) represents untapped compute: our codebase includes precompiled ANE LoRA kernels that could run adapter forward passes on ANE while the GPU handles training---true hardware parallelism for concurrent multi-agent operation.

\subsection{From Conversation to Persistent Specialization}

The ultimate vision is agents that develop persistent expertise through conversation:

\begin{enumerate}[noitemsep]
    \item \textbf{Session learning}: Agent absorbs facts from current conversation (demonstrated)
    \item \textbf{Cross-session persistence}: Adapter weights saved and restored across sessions (implemented, via \texttt{/save} endpoint)
    \item \textbf{Curriculum learning}: Progressive fact introduction with spaced repetition, using the replay buffer to revisit older knowledge
    \item \textbf{Collaborative specialization}: Different agents develop complementary expertise, forming a ``cognitive swarm'' that is collectively more capable than any single model
\end{enumerate}

\section{Reproduction}

\subsection{Prerequisites}
\begin{lstlisting}[language=bash, numbers=none]
pip install mlx mlx-lm fastapi uvicorn
\end{lstlisting}

\subsection{Self-Test}
\begin{lstlisting}[language=bash, numbers=none]
python3 mlx_lora_trainer.py
# Downloads Qwen2.5-0.5B-Instruct, trains 5 steps, verifies
\end{lstlisting}

\subsection{Full E2E via JARVIS Daemon}
\begin{lstlisting}[language=bash, numbers=none]
# Terminal 1
python3 neural_daemon.py

# Terminal 2
curl -X POST http://localhost:8766/activate \
  -d '{"hf_repo":"Qwen/Qwen3.5-2B-Base"}'
python3 test_daemon_e2e.py    # 4 facts, 20s
python3 test_deep_e2e.py      # 41 facts, 121s
\end{lstlisting}

\section{Conclusion}

We have demonstrated that real-time knowledge injection into language models is practical on consumer hardware. A 2B model learns novel facts in 20 seconds with perfect recall, and scales to 41 facts across 10 domains with 69\% recall and zero catastrophic forgetting. The critical parameters are learning rate ($5 \times 10^{-4}$, 10$\times$ standard LoRA) and regularization ratio ($\geq$33\% real-world data).

This work validates the foundation for conversational AI agents that learn from interaction. The path from single-agent adaptation to swarm architectures is clear: composable LoRA adapters, asynchronous training, and ANE-GPU parallelism. The agent that remembers, corrects, and specializes through conversation is no longer theoretical---it runs on a MacBook in 20 seconds.

\bibliographystyle{plain}
\begin{thebibliography}{10}

\bibitem{hu2021lora}
E.~J. Hu et al.
\newblock LoRA: Low-rank adaptation of large language models.
\newblock {\em arXiv:2106.09685}, 2021.

\bibitem{mlx2023}
A.~Hannun et al.
\newblock MLX: An array framework for Apple Silicon.
\newblock {\em Apple ML Research}, 2023.

\bibitem{dettmers2023qlora}
T.~Dettmers et al.
\newblock QLoRA: Efficient finetuning of quantized language models.
\newblock {\em arXiv:2305.14314}, 2023.

\bibitem{mccloskey1989catastrophic}
M.~McCloskey and N.~J. Cohen.
\newblock Catastrophic interference in connectionist networks.
\newblock {\em Psych.\ of Learning and Motivation}, 24:109--165, 1989.

\bibitem{rolnick2019experience}
D.~Rolnick et al.
\newblock Experience replay for continual learning.
\newblock {\em NeurIPS}, 2019.

\bibitem{kirkpatrick2017overcoming}
J.~Kirkpatrick et al.
\newblock Overcoming catastrophic forgetting in neural networks.
\newblock {\em PNAS}, 114(13):3521--3526, 2017.

\bibitem{gu2023mamba}
A.~Gu and T.~Dao.
\newblock Mamba: Linear-time sequence modeling with selective state spaces.
\newblock {\em arXiv:2312.00752}, 2023.

\bibitem{yang2024gated}
S.~Yang et al.
\newblock Gated delta networks: Improving mamba2 with delta rule.
\newblock {\em arXiv:2412.06464}, 2024.

\end{thebibliography}

\appendix

\section{Complete Training Data: Simple Test}
\label{app:simple}

The simple E2E test uses 11 training pairs (8 novel + 3 regularization). Each novel fact has 2--3 phrasing variants to improve generalization. See \texttt{test\_daemon\_e2e.py} for the full dataset.

\section{Complete Training Data: Deep Test}
\label{app:deep}

The deep E2E test uses 61 training pairs (41 novel across 10 domains + 20 regularization). Domains are interconnected with 15+ cross-references enabling multi-hop reasoning. See \texttt{test\_deep\_e2e.py} for the full dataset and 70-case test battery.

\section{Hyperparameter Sensitivity}
\label{app:sensitivity}

\begin{table}[h]
\centering
\caption{Full parameter sensitivity analysis}
\begin{tabular}{llcl}
\toprule
\textbf{Parameter} & \textbf{Range Tested} & \textbf{Optimal} & \textbf{Sensitivity} \\
\midrule
Learning rate & $5\text{e-5}$ -- $5\text{e-4}$ & $5\text{e-4}$ & \textbf{High} --- 10$\times$ effect on speed \\
Regularization ratio & 16\% -- 50\% & $\geq$33\% & \textbf{High} --- below 33\% causes forgetting \\
LoRA rank & 8 -- 32 & 32 & Low --- no speed difference \\
Early stop threshold & 0.1 -- 0.8 & 0.8 & Medium --- too low = wasted epochs \\
Gradient clip & 0.5 -- 2.0 & 1.0 & Low --- standard value works \\
Batch size & 1 -- 8 & 1 & None --- batching hurts on Apple Silicon \\
\texttt{mx.compile()} & on/off & off & \textbf{Negative} --- adds 20s overhead \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
